---
title: "EWD840"
author: "Markus A Kuppe"
date: "`r Sys.Date()`"
output: 
  github_document:
    toc: true
    toc_depth: 2
    html_preview: false
---

#### This file is generated on a host with all libraries installed via:
```shell
Rscript -e "rmarkdown::render('README.Rmd')"
```

## What

As part of the preparations to discuss Dijkstra's termination detection algorithm ([EWD840](https://www.cs.utexas.edu/users/EWD/ewd08xx/EWD840.PDF)) in a new [Dr. TLA+](https://github.com/tlaplus/DrTLAPlus) talk, we simulated the [TLA+ model](./EWD840.tla) with TLC to obtain statistics.  The basic idea was to generate as many behaviors with TLC's simulation mode as possible and see how spec-level changes influence the average length of behaviors until the algorithm detects termination.  The spec-level changes studied are:

 a) Color a node black regardless of where it sends an activation message (`SendMsg!SM1`)
 b) Pass along a black token independently of the node's color (`PassToken!PT1`)
 c) Pass along the token even if the node is active (`PassToken!PT2`)
 
For the formal definition of a, b, c, please see the labels `SM1`, `PT1`, and `PT2` in [EWD840.tla](./EWD840.tla).

### Correctness

Before we collected statistics, we verified the invariants `Inv` and `IInv` for all feature combinations and constants `N \in 1..7` with both TLC and [Apalache](https://github.com/informalsystems/apalache) (version `0.11.1-SNAPSHOT build v0.11.0-24-g293c4b2`). Note that Apalache required a [minor modification of Dijkstra's inductive invariant](https://github.com/informalsystems/apalache/issues/446) (see [apalache.md](./apalache.md)):

```tla
I ==
    {i \in 0..N-1: 0 <= i /\ i <= tpos}

\* Dijkstra's (inductive) invariant.
IInv ==
    \/ P0:: \A i \in Node : tpos < i => ~ active[i]
    \/ P1:: \E j \in I : color[j] = black
    \/ P2:: tcolor = black

```

Side note: This experiment surfaced a few issues in Apalache:

  * https://github.com/informalsystems/apalache/issues/654
  * https://github.com/informalsystems/apalache/issues/653
  * https://github.com/informalsystems/apalache/issues/425


### Data Collection

To minimize potential sources of error, we pushed data collection into the TLA+ specification (also see module [`EWD840_sim`](./EWD840_sim.tla)) and added two minor features to TLC. The first feature was to add support to simulation mode to evaluate a constant-level expression after a behavior is generated.  With this, we can write and flush per-behavior statistics that are stored in `TLCGet/TLCSet` registers:

```tla
---- MODULE EWD840_sim ----
...
PostBehavior == 
    /\ LET l == TLCGet("level")
           ip == TLCGet(1)
           sm == TLCGet(2)
           pt == TLCGet(3)
           da == TLCGet(4)
           ft == feature
           IO == INSTANCE IOUtils
        IN IO!IOCSV("%1$s#%2$s#%3$s#%4$s#%5$s#%6$s#%7$s", <<ft, l, ip, sm, pt, da, N>>, "out.csv")
    /\ \A i \in 1..4: TLCSet(i, 0)

=====
```

Secondly, we added a `CSV` operator (with a Java module override) to the [TLA+ CommunityModules](https://github.com/tlaplus/CommunityModules) to efficiently write the `TLCGet/TLCSet` registers to a file on disk:

```tla
CSV!CSVWrite(template, data, file)
```

Note that both TLC features have not yet been made available publicly.

## Results

Without further ado, below are plots of the average behavior length for varying numbers of nodes (`N=7`, `N=42`, `N=230`), depending on which one of the three "optimizations" `SM1`, `PT1`, and `PT2` are enabled:

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### N=7

```{r dataN7, echo=FALSE, warning=FALSE, message=FALSE}
library(here)
data <- read.csv(header=TRUE, sep = "#", file = here("N7.csv"))
```

```{r summaryN7, echo=TRUE, warning=FALSE, message=FALSE}
## R script
library(dplyr)
by_variant <- group_by(data,Variant)
summary = summarise(by_variant,
                    mean_length = mean(Length),
                    var_length = var(Length),
                    sd_length = sd(Length),
                    mean(InitiateProbe),
                    mean(SendMsg),
                    mean(PassToken),
                    mean(Deactivate),
                    N = mean(Node),
                    T = mean(Length-T)
)
```


```{r summaryTableN7}
library(knitr)
kable(format = "markdown", summary[, c(1,2,5,6,7,8,10)], digits=2, caption = "Data N=7 grouped by 'Variant'")
```

The summary confirms that the number of `Deactivate`, `PassToken`, `SendMsg` and `InitiateProbe` actions is the same regardless of the enabled optimizations.


```{r plotN7, echo=TRUE, warning=FALSE, message=FALSE}
## R script
library(ggplot2)
ggplot(summary, 
  aes(x = reorder(Variant, mean_length), y = mean_length, fill = Variant)) +
  geom_bar(stat = "identity") +
  geom_errorbar(aes(ymin=mean_length-sd_length, ymax=mean_length+sd_length), width=.2,
                position=position_dodge(.9)) +
  theme_minimal() +
  labs(
    x = "Spec variant",
    y = "Average length of behaviors",
    title = paste(
      "Number of Nodes:", data[,7][1], "Traces:", nrow(data)
  )
)
```

### N=42

```{r dataN42, echo=FALSE, warning=FALSE, message=FALSE}
library(here)
data <- read.csv(header=TRUE, sep = "#", file = here("N42.csv"))
```

```{r plotN42, echo=FALSE, warning=FALSE, message=FALSE}
library(dplyr)
by_variant <- group_by(data,Variant)
summary = summarise(by_variant,
                    mean_length = mean(Length),
                    var_length = var(Length),
                    sd_length = sd(Length),
                    mean(InitiateProbe),
                    mean(SendMsg),
                    mean(PassToken),
                    mean(Deactivate),
                    N = mean(Node),
                    T = mean(Length-T)
)
library(ggplot2)
ggplot(summary, 
  aes(x = reorder(Variant, mean_length), y = mean_length, fill = Variant)) +
  geom_bar(stat = "identity") +
  geom_errorbar(aes(ymin=mean_length-sd_length, ymax=mean_length+sd_length), width=.2,
                position=position_dodge(.9)) +
  theme_minimal() +
  labs(
    x = "Spec variant",
    y = "Average length of behaviors",
    title = paste(
      "Number of Nodes:", data[,7][1], "Traces:", nrow(data)
  )
)
```

### N=230

```{r dataN230, echo=FALSE, warning=FALSE, message=FALSE}
library(here)
data <- read.csv(header=TRUE, sep = "#", file = here("N230.csv"))
```

```{r plotN230, echo=FALSE, warning=FALSE, message=FALSE}
library(dplyr)
by_variant <- group_by(data,Variant)
summary = summarise(by_variant,
                    mean_length = mean(Length),
                    var_length = var(Length),
                    sd_length = sd(Length),
                    mean(InitiateProbe),
                    mean(SendMsg),
                    mean(PassToken),
                    mean(Deactivate),
                    N = mean(Node),
                    T = mean(Length-T)
)
library(ggplot2)
ggplot(summary, 
  aes(x = reorder(Variant, mean_length), y = mean_length, fill = Variant)) +
  geom_bar(stat = "identity") +
  geom_errorbar(aes(ymin=mean_length-sd_length, ymax=mean_length+sd_length), width=.2,
                position=position_dodge(.9)) +
  theme_minimal() +
  labs(
    x = "Spec variant",
    y = "Average length of behaviors",
    title = paste(
      "Number of Nodes:", data[,7][1], "Traces:", nrow(data)
  )
)
```

Statistically, the "optimization" `SM1` correlates with the shortest average behavior length, which follows the intuition that tainting the node if it activates lower-numbered nodes, results in additional rounds before termination is detect.  Surprisingly though, `PT1` and `PT2` have negative effects on the average behavior lenght--it takes longer to detect termination.

------

## Random/Eye-ball statistics

### Data distributions (behavior length and SendMsg actions) for N=7

```{r dataDistributionN7, echo=TRUE, warning=FALSE, message=FALSE}
library(ggplot2)
ggplot(data = data, mapping = aes(x=Length)) + 
  geom_histogram(aes(y=..density..),fill="bisque",color="white",alpha=0.9) + 
  geom_density() +
  geom_rug() +
  labs(x='Trace Length') +
  theme_minimal()

ggplot(data = data, mapping = aes(x=SendMsg)) + 
  geom_histogram(aes(y=..density..),fill="bisque",color="white",alpha=0.9) + 
  geom_density() +
  geom_rug() +
  labs(x='SendMsg') +
  theme_minimal()
```

### Correlation matrix for N=7

```{r dataCorrelationN7, echo=TRUE, warning=FALSE, message=FALSE}
##install.packages("ggcorrplot")
library("ggcorrplot")
my_data <- data[, c("Length", "SendMsg", "InitiateProbe", "PassToken", "Deactivate")]
p.mat <- cor_pmat(my_data)

## Check for correlation in 'data'
## 'spearman' (3) correlation because data has no normal distribution
## (see previous plots).
corr <- round(cor(my_data), 3)

ggcorrplot(corr, p.mat = cor_pmat(my_data),
           hc.order = TRUE, type = "lower",
           color = c("#FC4E07", "white", "#00AFBB"),
           outline.col = "white", lab = TRUE)
```

The correlation of TLA+ actions pretty much follows causality.

